\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-11-13 response.tex

Likewise, in Fig. 5, the negative correlation between drift in different clades in each season could, again, be driven by variation in the number of strains per clade that were subjected to HI each season. The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.

I found the section fitting diffusion coefficients (in different branches) to the antigenic maps confusing -- if only because strict diffusion (in one or two dimensions) has no advection term and therefore no tendency to move in one direction or another; whereas the data clearly show an advective tendency (moving to the ``right'' over time in antigenic dimension 1). I presume the authors attempt to deal with this by imposing a prior (eq 12) for the location of a virus based on its location in a molecular phylogeny. But this would seem to conflate antigenic drift, which the authors wish to quantify using a diffusion coefficient, with genetic mutations (many of which are driven by selection for antigenic escape). In other words, this hybrid approach of assuming a prior based on a strain's genotype, and then fitting a diffusion coefficient to summarize strength of drift, does not seem to to be a pure measure of drift. What exactly the fitted diffusion coefficient means is extremely unclear to me, in any formal sense, as the procedure is so ad-hoc.

\textbf{We agree that applying the standard measure of diffusion coefficient $D$ based on displacement vs time, without accounting for advection was problematic.  We have substantial revised the text to focus on direct parameter inference of drift parameter $\drift$ and diffusion volatility parameter $\virussd$, properly separating advection and volatility in the diffusion process.  The section ``Antigenic evolution across influenza lineages'' most reflects these changes, including the addition of a new table 2.}

### 2013-11-13 response.tex

IMPORTANT BUT NOT SCIENTIFICALLY SUBSTANTIVE POINTS:

The authors should be commended for making publicly available all of the HI data that they have assembled (a good scientific practice that sadly has not been adhered to in some previous publications on antigenic cartography). However, they should mention in the text the availability of these data on datadryad - I almost missed it since it is only mentioned on the cover page. They could also consider more clearly specifying where the code relevant to this paper is found in the BEAST repository.

The one exception to the aforementioned praise on data availability is the influenza incidence data - I could not find any tables providing these data. They need to be provided.

\textbf{We apologize for this oversight.  The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.  We've also included a reference to the BEAST tutorials on implementing these cartographic models.}

### 2013-11-14 flux.tex
Work on issue #4.
Text and response is finished, however need to include MDS tutorial
on the BEAST wiki.
Significant revision of drift and incidence analysis per #1 and #2.

We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}).
We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.

%%% jumps %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=1.0\textwidth]{figures/jumps}
	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2009 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011 and seasonal incidence from 1998/1999 to 2008/2009.
	Colored lines represent year-to-year antigenic drift, and drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
	For example, 2000 represents difference in antigenic dimension 1 between 1999 and 2000.
	Error bars represent 50\% Bayesian credible intervals of year-to-year drift.
	Gray dotted lines represent clade-specific USA incidence taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a clade for each season.
	Here, 2000 represents the 2000/2001 influenza season.
	(B) Histograms of year-to-year antigenic drift across influenza clades summarizing observations from 1992 to 2011.
	} 
	\label{jumps} 
\end{figure}

We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Materials and methods).
We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of viruses detected in the USA during this time period.
We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).

We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of viruses attributable to a clade for each season.
This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{jumps}, Figure~\ref{incidence}A). 
Here, within-clade drift explains 35\% of the variance in standardized incidence and represents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).

%%% incidence %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/incidence}
	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
	} 
	\label{incidence} 
\end{figure}

We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
We fit a linear model that predicts standardized incidence based on drift within each of the four lineages, according to the following equations
\begin{eqnarray}
	\incclade{H3} &=& \bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
	\incclade{H1} &=& \bsister \, \driftclade{H3} + \bwithin \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
	\incclade{Vic} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bwithin \, \driftclade{Vic} + \bsister \, \driftclade{Yam} \nonumber \\	
	\incclade{Yam} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bsister \, \driftclade{Vic} + \bwithin \, \driftclade{Yam},
\end{eqnarray}
where $y$ represents standardized incidence, $x$ represents drift in antigenic units, $\bwithin$ scales within-clade drift, $\bsister$ scales drift across sister viruses of the same type, and $\bother$ scales drift across viruses of differing type (Table~\ref{interferencetable}).
For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2 scaled by $\bwithin$, drift in A/H1N1 scaled by $\bsister$ and by drift in B/Vic and B/Yam scaled by $\bother$.
We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations.
The term for across lineage interference $\bother$ is small in magnitude, but remains statistically significant (Table~\ref{interferencetable}).

%%% interferencetable %%%
\begin{table}[h]
	\centering
	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
	\label{interferencetable}	
	\begin{tabular}{ c c c c } 
	\hline
	Parameter	&	Estimate	&	Standard error 	& 	$p$-value \\
	\hline		
	$\bwithin$	&	$1.26$		&	0.22			&	$1.1 \times 10^{-6}$ \\
	$\bsister$	&	$-0.61$		&	0.22			&	$0.008$ \\
	$\bother$	&	$-0.32$		&	0.11			&	$0.005$ \\
	\hline
	\end{tabular}
\end{table}

These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence across  A/H3N2, A/H1N1, B/Vic and B/Yam is strongly influenced by levels of antigenic drift within each clade, consistent with antigenic drift increasing the susceptible pool available to a clade.
The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
The immunological basis of between-clade interference are not entirely clear, however cytotoxic T cells (CTLs) \cite{Voeten00}, antibodies that broadly neutralize HA \cite{Corti10} and antibodies to internal proteins \cite{Laidlaw13} have been hypothesized as possible mechanisms.
Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2 (Table~\ref{interferencetable}, $\bsister = -0.61$ vs $\bother = -0.32$).
If interference is mediated by CTLs or by antibodies to evolutionary conserved epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.

Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.

@@

Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
Incidence data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).

>>>>

We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}B), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.

%%% jumps %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=1.0\textwidth]{figures/jumps}
	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
	(A) Timeseries of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
	Colored lines represent year-to-year antigenic drift, where drift for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
	For example, 2000 represents difference in antigenic dimension 1 between viruses from 1999 and 2000.
	Error bars represent 50\% Bayesian credible intervals of year-to-year drift.
	Gray dotted lines represent lineage-specific seasonal incidence in the USA taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a lineage for each season.
	Here, 2000 represents the 2000/2001 influenza season.
	(B) Distribution of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
	} 
	\label{jumps} 
\end{figure}

We investigate the relationship between rates of antigenic drift and seasonal incidence in the USA in A/H3N2, A/H1N1, B/Vic and B/Yam.
We measure seasonal incidence from USA CDC data for each virus lineage (A/H3N2, A/H1N1, B/Vic, B/Yam) by taking the average influenza-like (ILI) percentage in a season and multiplying this by the relative proportion of virus isolations attributable to a particular influenza lineage in a season (see Materials and methods).
This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic.
We begin by comparing overall rates of antigenic drift (Figure~\ref{drift}, Table~\ref{paramtable}) to overall levels of seasonal incidence across influenza lineages, finding a significant correlation between rate of antigenic drift and relative incidence across the four lineages (Pearson correlation, $r=0.97$, $p = 0.041$).

We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and lineage-specific incidence, comparing incidence in a seasonal to antigenic drift of viruses coming into this season (Figure~\ref{jumps}A).
For example, we compare antigenic drift of viruses from 2000 to 2001 to incidence in the 2001/2002 season.
Within each virus lineage, we find that years with pronounced antigenic drift tend to show increased incidence (Figure~\ref{corr}), finding Pearson correlation coefficients of 0.51, 0.29, 0.44 and 0.14 for A/H3N2, A/H1N1, B/Vic and B/Yam respectively.
We calculated significance using bootstrap permutation tests finding $p$-values of 0.056, 0.201, 0.097 and 0.341 respectively.
We applied a similar bootstrap permutation test to calculate the significance of finding the observed levels of correlation across all four lineages, arriving at a $p$-value of 0.018.

%%% corr %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.7\textwidth]{figures/corr}
	\caption{\textbf{Relationship between antigenic drift and seasonal incidence for years 1998 to 2009 in influenza A/H3N2, A/H1N1, B/Vic and B/Yam.}
	Antigenic drift from year $i-1$ to year $i$ is compared to incidence in the season $i$ / $i+1$.
	For example, year-to-year antigenic drift from 2000 to 2001 is measured against incidence in the 2001/2002 season. 
	} 
	\label{corr} 
\end{figure}

Combining observations of year-to-year drift vs seasonal incidence across all four lineages, yields a simple linear model where one unit of antigenic drift results in an increase of 0.8\% in lineage-specific ILI.

%Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
%However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
%Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.

@@

Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
Tutorials for implementing these models can be found at http://beast.bio.ed.ac.uk/Tutorials.
Incidence data and HI data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
